---
title: "USA Airline Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document: default
---

## Reading and Exploring the Dataset

```{r setup, include=FALSE}
install.packages("zoo")
install.packages("ggcorrplot")
install.packages(c( "ggplot2","GGally"))
install.packages("tidyr")
```

```{r beginning}
# Load necessary library
library(dplyr)
library(zoo)
library(ggcorrplot)
library(tidyr)


# Reading the USA Airline data
airline_data <- read.csv("C:/Users/EWURA/Desktop/PROJECT-Professor Mashouri/PROJECT/Dataset/usa.csv", stringsAsFactors = FALSE)

# Display basic information about the dataset
str(airline_data)

```

```{r 1}
# Display the first 5 rows of the dataset
head(airline_data, 5)

# Check for missing values
missing_values <- sapply(airline_data, function(x) sum(is.na(x)))
missing_values[missing_values > 0]  # Show columns with missing values

```

## Handling Missing Values

###  Check for Missing Values
```{r check-missing-values}
# Check for missing values
missing_values <- sapply(airline_data, function(x) sum(is.na(x)))
missing_values[missing_values > 0]  # Display columns with missing values

# Impute missing values in carrier-related columns with "Unknown"
airline_data$carrier_lg <- ifelse(is.na(airline_data$carrier_lg), "Unknown", airline_data$carrier_lg)
airline_data$carrier_low <- ifelse(is.na(airline_data$carrier_low), "Unknown", airline_data$carrier_low)

# Replace missing numeric values with the column mean
airline_data$large_ms <- ifelse(is.na(airline_data$large_ms), mean(airline_data$large_ms, na.rm = TRUE), airline_data$large_ms)
airline_data$fare_lg <- ifelse(is.na(airline_data$fare_lg), mean(airline_data$fare_lg, na.rm = TRUE), airline_data$fare_lg)
airline_data$lf_ms <- ifelse(is.na(airline_data$lf_ms), mean(airline_data$lf_ms, na.rm = TRUE), airline_data$lf_ms)
airline_data$fare_low <- ifelse(is.na(airline_data$fare_low), mean(airline_data$fare_low, na.rm = TRUE), airline_data$fare_low)
 
# Impute missing values in Geocoded_City1 and Geocoded_City2
airline_data$Geocoded_City1 <- ifelse(is.na(airline_data$Geocoded_City1), airline_data$city1, airline_data$Geocoded_City1)
airline_data$Geocoded_City2 <- ifelse(is.na(airline_data$Geocoded_City2), airline_data$city2, airline_data$Geocoded_City2)
            
# Verify remaining missing values
remaining_missing_values <- sapply(airline_data, function(x) sum(is.na(x)))
remaining_missing_values
```
## Combining and Manipulating Data

### Combine `Year` and `quarter` to Create a Time Variable
```{r create-time-variable}
# Create a time variable by combining Year and quarter
airline_data$time <- paste(airline_data$Year, airline_data$quarter, sep = "-Q")
airline_data$time <- as.yearqtr(airline_data$time, format = "%Y-Q%q")  # Convert to year-quarter format

# Sort the dataset by the time variable
airline_data <- airline_data[order(airline_data$time), ]
head(airline_data[, c("time", "Year", "quarter")])

# Aggregate data by time, airport_1, and airport_2
library(dplyr)
route_specific_aggregation <- airline_data %>%
  group_by(time, airport_1, airport_2) %>%
  summarise(
    passengers = sum(passengers, na.rm = TRUE),  # Total passengers
    fare = mean(fare, na.rm = TRUE),            # Average fare
    nsmiles = mean(nsmiles, na.rm = TRUE)       # Average distance
  ) %>%
  ungroup()

# Display the aggregated data
print("Route-Specific Aggregated Data:")
head(route_specific_aggregation)

# Count the number of arrivals for each destination airport
arrivals <- airline_data %>%
  group_by(airportid_2) %>%
  summarise(num_arrivals = n())

# Merge arrivals data back into the original dataset
airline_data <- left_join(airline_data, arrivals, by = "airportid_2")

# Split city2 into city and state
library(stringr)
city_state <- str_split_fixed(airline_data$city2, ",", 2)
airline_data$city <- str_trim(city_state[, 1])  # Extract and trim city
airline_data$state <- str_trim(city_state[, 2]) # Extract and trim state

# Check the updated dataset
head(airline_data[, c("city2", "city", "state")])

# Check and remove duplicate columns, if any
if ("num_arrivals_x" %in% names(airline_data) & "num_arrivals_y" %in% names(airline_data)) {
  # Consolidate the columns (assuming they are identical)
  airline_data$num_arrivals <- airline_data$num_arrivals_y
  airline_data <- airline_data %>% select(-num_arrivals_x, -num_arrivals_y)
}

# Verify the result
head(airline_data[, c("airportid_2", "num_arrivals")])
```
## Cleaning the Dataset

### Drop Unnecessary Columns
```{r drop-columns}
# Specify columns to drop
columns_to_drop <- c(
  "city1", "city2", "carrier_lg", "large_ms", "fare_lg",
  "carrier_low", "lf_ms", "fare_low", "Geocoded_City1", "Geocoded_City2",
  "citymarketid_1", "citymarketid_2", "airportid_1", "airportid_2"
)

# Remove the columns from the dataset
airline_data_cleaned <- airline_data %>% select(-all_of(columns_to_drop))

# Display the remaining columns
colnames(airline_data_cleaned)

# Preview the cleaned dataset
head(airline_data_cleaned)
```
## Summary Statistics and Visualizations

### Summary Statistics for the Cleaned Dataset
```{r summary-statistics}
# Summary statistics for the cleaned dataset
summary(airline_data_cleaned)

# Count rows with zero passengers
zero_passengers <- airline_data_cleaned %>% filter(passengers == 0)
cat("Number of rows with zero passengers:", nrow(zero_passengers), "\n")

# Display some examples
cat("Examples of rows with zero passengers:")
head(zero_passengers)
```

```{r Aggeregation}
# Aggregate total passengers over time
library(ggplot2)
time_trend <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(total_passengers = sum(passengers, na.rm = TRUE)) %>%
  ungroup()

# Plot passenger trends over time
ggplot(time_trend, aes(x = time, y = total_passengers)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue") +
  labs(
    title = "Total Passengers Over Time",
    x = "Time",
    y = "Total Passengers"
  ) +
  theme_minimal()

# Aggregate average fare over time
fare_trend <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_fare = mean(fare, na.rm = TRUE)) %>%
  ungroup()

# Plot average fare trends over time
ggplot(fare_trend, aes(x = time, y = average_fare)) +
  geom_line(color = "orange", size = 1) +
  geom_point(color = "orange") +
  labs(
    title = "Average Fare Over Time",
    x = "Time",
    y = "Average Fare"
  ) +
  theme_minimal()

# Aggregate total passengers by route
top_routes <- airline_data_cleaned %>%
  group_by(airport_1, airport_2) %>%
  summarise(total_passengers = sum(passengers, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(total_passengers)) %>%
  head(10)

cat("Top 10 Routes by Total Passengers:")
print(top_routes)

# Aggregate average fares by route
high_fare_routes <- airline_data_cleaned %>%
  group_by(airport_1, airport_2) %>%
  summarise(average_fare = mean(fare, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(average_fare)) %>%
  head(10)

cat("Top 10 Routes by Average Fare:")
print(high_fare_routes)

```
## Analysis and Visualization

### Aggregate Passengers by Quarter (Seasonality)
```{r seasonality, warning=FALSE, message=FALSE}
# Aggregate average passengers by quarter
library(dplyr)
seasonality <- airline_data_cleaned %>%
  mutate(Quarter = as.numeric(format(time, "%q"))) %>%  # Extract quarter from time
  group_by(Quarter) %>%
  summarise(Average_Passengers = mean(passengers, na.rm = TRUE)) %>%
  ungroup()

# Plot seasonality
library(ggplot2)
ggplot(seasonality, aes(x = Quarter, y = Average_Passengers)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average Passengers by Quarter",
    x = "Quarter",
    y = "Average Passengers"
  ) +
  theme_minimal()

# Compute correlation matrix for numerical columns
numeric_columns <- airline_data_cleaned %>%
  select(passengers, fare, nsmiles)
correlation_matrix <- cor(numeric_columns, use = "complete.obs")

# Print correlation matrix
print("Correlation Matrix:")
print(correlation_matrix)

# Visualize the correlation matrix
library(ggcorrplot)
ggcorrplot(correlation_matrix, lab = TRUE, colors = c("red", "white", "blue")) +
  labs(title = "Correlation Matrix")

# Plot distribution of passengers
ggplot(airline_data_cleaned, aes(x = passengers)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  labs(
    title = "Passenger Distribution",
    x = "Passengers",
    y = "Frequency"
  ) +
  theme_minimal()

# Plot distribution of fares
ggplot(airline_data_cleaned, aes(x = fare)) +
  geom_histogram(bins = 50, fill = "orange", color = "black") +
  labs(
    title = "Fare Distribution",
    x = "Fare",
    y = "Frequency"
  ) +
  theme_minimal()

# Apply log transformation to passengers
airline_data_cleaned$log_passengers <- log1p(airline_data_cleaned$passengers)  # log1p handles zero values

# Plot the distribution of log-transformed passengers
ggplot(airline_data_cleaned, aes(x = log_passengers)) +
  geom_histogram(bins = 30, fill = "green", color = "black") +
  labs(
    title = "Distribution of Log-Transformed Passengers",
    x = "Log-Transformed Passengers",
    y = "Frequency"
  ) +
  theme_minimal()

# Apply log transformation to fares
airline_data_cleaned$log_fare <- log1p(airline_data_cleaned$fare)

# Plot the distribution of log-transformed fares
ggplot(airline_data_cleaned, aes(x = log_fare)) +
  geom_histogram(bins = 30, fill = "orange", color = "black") +
  labs(
    title = "Distribution of Log-Transformed Fares",
    x = "Log-Transformed Fare",
    y = "Frequency"
  ) +
  theme_minimal()

```
## Boxplots for Log-Transformed Variables

### Boxplot for Log-Transformed Passengers
```{r boxplot-log-passengers, warning=FALSE, message=FALSE}
# Boxplot for log-transformed passengers
library(ggplot2)

ggplot(airline_data_cleaned, aes(x = log_passengers)) +
  geom_boxplot(fill = "green", color = "black") +
  labs(
    title = "Boxplot of Log-Transformed Passengers",
    x = "Log-Transformed Passengers"
  ) +
  theme_minimal()

# Boxplot for log-transformed fares
ggplot(airline_data_cleaned, aes(x = log_fare)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(
    title = "Boxplot of Log-Transformed Fares",
    x = "Log-Transformed Fares"
  ) +
  theme_minimal()
```
## Aggregations, Correlations, and Pairwise Relationships

### 1. Aggregate and Plot Average Log Passengers Over Time
```{r log-passengers-over-time, warning=FALSE, message=FALSE}
# Aggregate log_passengers by time
library(dplyr)
library(ggplot2)

average_passengers_over_time <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_log_passengers = mean(log_passengers, na.rm = TRUE)) %>%
  ungroup()

# Plot the aggregated time series
ggplot(average_passengers_over_time, aes(x = time, y = average_log_passengers)) +
  geom_line(color = "green", size = 1) +
  labs(
    title = "Average Log Passengers Over Time",
    x = "Time",
    y = "Average Log Passengers"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey90"))

# Aggregate log_fare by time
average_fare_over_time <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_log_fare = mean(log_fare, na.rm = TRUE)) %>%
  ungroup()

# Plot the aggregated time series
ggplot(average_fare_over_time, aes(x = time, y = average_log_fare)) +
  geom_line(color = "orange", size = 1) +
  labs(
    title = "Average Log Fare Over Time",
    x = "Time",
    y = "Average Log Fare"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey90"))

# Aggregate log_fare by time
average_fare_over_time <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_log_fare = mean(log_fare, na.rm = TRUE)) %>%
  ungroup()

# Plot the aggregated time series
ggplot(average_fare_over_time, aes(x = time, y = average_log_fare)) +
  geom_line(color = "orange", size = 1) +
  labs(
    title = "Average Log Fare Over Time",
    x = "Time",
    y = "Average Log Fare"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey90"))

# Create pairwise scatterplots
library(GGally)

ggpairs(
  airline_data_cleaned,
  columns = c("log_fare", "log_passengers", "nsmiles"),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "smooth"),
  upper = list(continuous = "cor")
) +
  labs(title = "Pairwise Relationships Among Key Variables") +
  theme_minimal()

```
```{r state_trend_plot, echo=TRUE, message=FALSE, warning=FALSE}

state_trend <- airline_data_cleaned %>%
  group_by(time, state) %>%
  summarise(total_arrivals = sum(num_arrivals, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = state, values_from = total_arrivals)

# Identify the top 5 states by total arrivals
top_states <- colSums(state_trend[-1], na.rm = TRUE) %>%
  sort(decreasing = TRUE) %>%
  head(5) %>%
  names()

# Filter and reshape data for plotting
plot_data <- state_trend %>%
  select(time, all_of(top_states)) %>%
  gather(key = "state", value = "total_arrivals", -time)

# Plot the trend for the top 5 states
ggplot(plot_data, aes(x = time, y = total_arrivals, color = state, group = state)) +
  geom_line(size = 1.2, marker = "o") +
  labs(title = "Trend of Arrivals for Top 5 States", x = "Time", y = "Total Arrivals") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  scale_color_discrete(name = "State")

```
```{r city_trend_plot, echo=TRUE, message=FALSE, warning=FALSE}

# Group by time and city, and calculate total arrivals
city_trend <- airline_data_cleaned %>%
  group_by(time, city) %>%
  summarise(total_arrivals = sum(num_arrivals, na.rm = TRUE)) %>%
  spread(key = city, value = total_arrivals)

# Identify the top 5 cities by total arrivals
top_cities <- colSums(city_trend[-1], na.rm = TRUE) %>%
  sort(decreasing = TRUE) %>%
  head(5) %>%
  names()

# Filter and reshape data for plotting
plot_data <- city_trend %>%
  select(time, all_of(top_cities)) %>%
  gather(key = "city", value = "total_arrivals", -time)

# Plot the trend for the top 5 cities
ggplot(plot_data, aes(x = time, y = total_arrivals, color = city, group = city)) +
  geom_line(size = 1.2, marker = "o") +
  labs(title = "Trend of Arrivals for Top 5 Cities", x = "Time", y = "Total Arrivals") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    legend.title = element_text(size = 12),
    legend.text = element_text(size = 10)
  ) +
  scale_color_discrete(name = "City")

```

```{r inflation_adjustment, echo=TRUE, message=FALSE, warning=FALSE}

# Set the inflation rate
inflation_rate <- 0.02  # 2% inflation

# Identify the base year
base_year <- max(airline_data_cleaned$Year)

# Adjust fares for inflation
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(adjusted_fare = fare * (1 + inflation_rate)^(base_year - Year))

# Calculate percentiles for the adjusted fares
adjusted_fare_percentiles <- quantile(airline_data_cleaned$adjusted_fare, probs = c(0.2, 0.4, 0.6, 0.8))

# Categorize the adjusted fares
categorize_adjusted_fare <- function(fare) {
  if (fare < adjusted_fare_percentiles[1]) {
    return("Very Low")
  } else if (fare >= adjusted_fare_percentiles[1] && fare < adjusted_fare_percentiles[2]) {
    return("Low")
  } else if (fare >= adjusted_fare_percentiles[2] && fare < adjusted_fare_percentiles[3]) {
    return("Medium")
  } else if (fare >= adjusted_fare_percentiles[3] && fare < adjusted_fare_percentiles[4]) {
    return("High")
  } else {
    return("Very High")
  }
}

# Apply the categorization to the adjusted fare
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(fare_category = sapply(adjusted_fare, categorize_adjusted_fare))

# Display the updated DataFrame
head(airline_data_cleaned %>% select(Year, fare, adjusted_fare, fare_category))
```

```{r hierarchy_code_creation, echo=TRUE, message=FALSE, warning=FALSE}


# Create the `time_point` column
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(time_point = paste0(Year, "Q", quarter))

# Define fare category and city mappings
fare_category_map <- c(
  "Very Low" = "VL",
  "Low" = "L",
  "Medium" = "M",
  "High" = "H",
  "Very High" = "VH"
)

city_short_map <- c(
  "San Francisco" = "SF",
  "New York" = "NY",
  "Los Angeles" = "LA",
  "Chicago" = "CH",
  "Washington" = "DC"
)

# Create the `fare_category_code` column by mapping categories
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(fare_category_code = fare_category_map[fare_category])

# Create the `city_short` column using the city map and fallback to first two letters if not mapped
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(city_short = ifelse(city %in% names(city_short_map),
                             city_short_map[city],
                             str_to_upper(substr(city, 1, 2))))

# Combine elements into a `hierarchy_code`
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(hierarchy_code = paste0(time_point, state, city_short, fare_category_code))

# Select the final columns for export
final_hierarchy_table <- airline_data_cleaned %>%
  select(time_point, state, city_short, fare_category_code, hierarchy_code, passengers, adjusted_fare)

# Display the first few rows of the final table
head(final_hierarchy_table)
```

```{r clean_state_and_hierarchy, echo=TRUE, message=FALSE, warning=FALSE}

# Load necessary libraries
library(stringr)

# Clean the `state` column to keep only the first two uppercase letters
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(state = ifelse(str_detect(state, "^[A-Z]{2}"), str_sub(state, 1, 2), state))

# Recreate the hierarchy code with the cleaned `state` column
airline_data_cleaned <- airline_data_cleaned %>%
  mutate(hierarchy_code = paste0(time_point, state, city_short, fare_category_code))

# Verify the updated hierarchy code column
cat("Updated Hierarchy Code Preview:\n")
print(head(airline_data_cleaned %>% select(hierarchy_code, state, city_short)))

# Select the final columns for export
final_hierarchy_table <- airline_data_cleaned %>%
  select(time_point, state, city_short, fare_category_code, hierarchy_code, passengers, adjusted_fare)

# Save the updated table to a CSV file (optional)
# write.csv(final_hierarchy_table, "final_hierarchy_table.csv", row.names = FALSE)

```

```{r export_final_hierarchy_table, echo=TRUE, message=FALSE, warning=FALSE}

# Load necessary libraries
library(dplyr)

# Select the final columns
final_hierarchy_table <- airline_data_cleaned %>%
  select(
    time_point,
    state,
    city_short,
    fare_category_code,
    hierarchy_code,
    passengers,
    adjusted_fare
  )

# Rename columns
final_hierarchy_table <- final_hierarchy_table %>%
  rename(
    timepoints = time_point,
    state = state,                  # No change for the `state` column name
    city = city_short,
    farecategory = fare_category_code,
    hierarchy = hierarchy_code
  )

final_hierarchy_table
```


