---
title: "USA Airline Data Analysis"
author: "Your Name"
date: "`r Sys.Date()`"
output: 
  html_document: default
  ipynb_notebook: default
---

## Reading and Exploring the Dataset

```{r setup, include=FALSE}
install.packages("zoo")
install.packages("ggcorrplot")
install.packages(c( "ggplot2","GGally"))
install.packages("rmarkdown")
install.packages("IRkernel")

# Load necessary library
library(dplyr)
library(zoo)
library(ggcorrplot)

# Reading the USA Airline data
airline_data <- read.csv("C:/Users/EWURA/Desktop/PROJECT-Professor Mashouri/PROJECT/Dataset/usa.csv", stringsAsFactors = FALSE)

# Display basic information about the dataset
str(airline_data)

```

```{r 1}
# Display the first 5 rows of the dataset
head(airline_data, 5)

# Check for missing values
missing_values <- sapply(airline_data, function(x) sum(is.na(x)))
missing_values[missing_values > 0]  # Show columns with missing values

```

## Handling Missing Values

###  Check for Missing Values
```{r check-missing-values}
# Check for missing values
missing_values <- sapply(airline_data, function(x) sum(is.na(x)))
missing_values[missing_values > 0]  # Display columns with missing values

# Impute missing values in carrier-related columns with "Unknown"
airline_data$carrier_lg <- ifelse(is.na(airline_data$carrier_lg), "Unknown", airline_data$carrier_lg)
airline_data$carrier_low <- ifelse(is.na(airline_data$carrier_low), "Unknown", airline_data$carrier_low)

# Replace missing numeric values with the column mean
airline_data$large_ms <- ifelse(is.na(airline_data$large_ms), mean(airline_data$large_ms, na.rm = TRUE), airline_data$large_ms)
airline_data$fare_lg <- ifelse(is.na(airline_data$fare_lg), mean(airline_data$fare_lg, na.rm = TRUE), airline_data$fare_lg)
airline_data$lf_ms <- ifelse(is.na(airline_data$lf_ms), mean(airline_data$lf_ms, na.rm = TRUE), airline_data$lf_ms)
airline_data$fare_low <- ifelse(is.na(airline_data$fare_low), mean(airline_data$fare_low, na.rm = TRUE), airline_data$fare_low)
 
# Impute missing values in Geocoded_City1 and Geocoded_City2
airline_data$Geocoded_City1 <- ifelse(is.na(airline_data$Geocoded_City1), airline_data$city1, airline_data$Geocoded_City1)
airline_data$Geocoded_City2 <- ifelse(is.na(airline_data$Geocoded_City2), airline_data$city2, airline_data$Geocoded_City2)
            
# Verify remaining missing values
remaining_missing_values <- sapply(airline_data, function(x) sum(is.na(x)))
remaining_missing_values
```
## Combining and Manipulating Data

### Combine `Year` and `quarter` to Create a Time Variable
```{r create-time-variable}
# Create a time variable by combining Year and quarter
airline_data$time <- paste(airline_data$Year, airline_data$quarter, sep = "-Q")
airline_data$time <- as.yearqtr(airline_data$time, format = "%Y-Q%q")  # Convert to year-quarter format

# Sort the dataset by the time variable
airline_data <- airline_data[order(airline_data$time), ]
head(airline_data[, c("time", "Year", "quarter")])

# Aggregate data by time, airport_1, and airport_2
library(dplyr)
route_specific_aggregation <- airline_data %>%
  group_by(time, airport_1, airport_2) %>%
  summarise(
    passengers = sum(passengers, na.rm = TRUE),  # Total passengers
    fare = mean(fare, na.rm = TRUE),            # Average fare
    nsmiles = mean(nsmiles, na.rm = TRUE)       # Average distance
  ) %>%
  ungroup()

# Display the aggregated data
print("Route-Specific Aggregated Data:")
head(route_specific_aggregation)

# Count the number of arrivals for each destination airport
arrivals <- airline_data %>%
  group_by(airportid_2) %>%
  summarise(num_arrivals = n())

# Merge arrivals data back into the original dataset
airline_data <- left_join(airline_data, arrivals, by = "airportid_2")

# Split city2 into city and state
library(stringr)
city_state <- str_split_fixed(airline_data$city2, ",", 2)
airline_data$city <- str_trim(city_state[, 1])  # Extract and trim city
airline_data$state <- str_trim(city_state[, 2]) # Extract and trim state

# Check the updated dataset
head(airline_data[, c("city2", "city", "state")])

# Check and remove duplicate columns, if any
if ("num_arrivals_x" %in% names(airline_data) & "num_arrivals_y" %in% names(airline_data)) {
  # Consolidate the columns (assuming they are identical)
  airline_data$num_arrivals <- airline_data$num_arrivals_y
  airline_data <- airline_data %>% select(-num_arrivals_x, -num_arrivals_y)
}

# Verify the result
head(airline_data[, c("airportid_2", "num_arrivals")])
```
## Cleaning the Dataset

### Drop Unnecessary Columns
```{r drop-columns}
# Specify columns to drop
columns_to_drop <- c(
  "city1", "city2", "carrier_lg", "large_ms", "fare_lg",
  "carrier_low", "lf_ms", "fare_low", "Geocoded_City1", "Geocoded_City2",
  "citymarketid_1", "citymarketid_2", "airportid_1", "airportid_2"
)

# Remove the columns from the dataset
airline_data_cleaned <- airline_data %>% select(-all_of(columns_to_drop))

# Display the remaining columns
colnames(airline_data_cleaned)

# Preview the cleaned dataset
head(airline_data_cleaned)
```
## Summary Statistics and Visualizations

### Summary Statistics for the Cleaned Dataset
```{r summary-statistics}
# Summary statistics for the cleaned dataset
summary(airline_data_cleaned)

# Count rows with zero passengers
zero_passengers <- airline_data_cleaned %>% filter(passengers == 0)
cat("Number of rows with zero passengers:", nrow(zero_passengers), "\n")

# Display some examples
cat("Examples of rows with zero passengers:")
head(zero_passengers)
```

```{r Aggeregation}
# Aggregate total passengers over time
library(ggplot2)
time_trend <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(total_passengers = sum(passengers, na.rm = TRUE)) %>%
  ungroup()

# Plot passenger trends over time
ggplot(time_trend, aes(x = time, y = total_passengers)) +
  geom_line(color = "blue", size = 1) +
  geom_point(color = "blue") +
  labs(
    title = "Total Passengers Over Time",
    x = "Time",
    y = "Total Passengers"
  ) +
  theme_minimal()

# Aggregate average fare over time
fare_trend <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_fare = mean(fare, na.rm = TRUE)) %>%
  ungroup()

# Plot average fare trends over time
ggplot(fare_trend, aes(x = time, y = average_fare)) +
  geom_line(color = "orange", size = 1) +
  geom_point(color = "orange") +
  labs(
    title = "Average Fare Over Time",
    x = "Time",
    y = "Average Fare"
  ) +
  theme_minimal()

# Aggregate total passengers by route
top_routes <- airline_data_cleaned %>%
  group_by(airport_1, airport_2) %>%
  summarise(total_passengers = sum(passengers, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(total_passengers)) %>%
  head(10)

cat("Top 10 Routes by Total Passengers:")
print(top_routes)

# Aggregate average fares by route
high_fare_routes <- airline_data_cleaned %>%
  group_by(airport_1, airport_2) %>%
  summarise(average_fare = mean(fare, na.rm = TRUE)) %>%
  ungroup() %>%
  arrange(desc(average_fare)) %>%
  head(10)

cat("Top 10 Routes by Average Fare:")
print(high_fare_routes)

```
## Analysis and Visualization

### Aggregate Passengers by Quarter (Seasonality)
```{r seasonality, warning=FALSE, message=FALSE}
# Aggregate average passengers by quarter
library(dplyr)
seasonality <- airline_data_cleaned %>%
  mutate(Quarter = as.numeric(format(time, "%q"))) %>%  # Extract quarter from time
  group_by(Quarter) %>%
  summarise(Average_Passengers = mean(passengers, na.rm = TRUE)) %>%
  ungroup()

# Plot seasonality
library(ggplot2)
ggplot(seasonality, aes(x = Quarter, y = Average_Passengers)) +
  geom_bar(stat = "identity", fill = "skyblue") +
  labs(
    title = "Average Passengers by Quarter",
    x = "Quarter",
    y = "Average Passengers"
  ) +
  theme_minimal()

# Compute correlation matrix for numerical columns
numeric_columns <- airline_data_cleaned %>%
  select(passengers, fare, nsmiles)
correlation_matrix <- cor(numeric_columns, use = "complete.obs")

# Print correlation matrix
print("Correlation Matrix:")
print(correlation_matrix)

# Visualize the correlation matrix
library(ggcorrplot)
ggcorrplot(correlation_matrix, lab = TRUE, colors = c("red", "white", "blue")) +
  labs(title = "Correlation Matrix")

# Plot distribution of passengers
ggplot(airline_data_cleaned, aes(x = passengers)) +
  geom_histogram(bins = 50, fill = "skyblue", color = "black") +
  labs(
    title = "Passenger Distribution",
    x = "Passengers",
    y = "Frequency"
  ) +
  theme_minimal()

# Plot distribution of fares
ggplot(airline_data_cleaned, aes(x = fare)) +
  geom_histogram(bins = 50, fill = "orange", color = "black") +
  labs(
    title = "Fare Distribution",
    x = "Fare",
    y = "Frequency"
  ) +
  theme_minimal()

# Apply log transformation to passengers
airline_data_cleaned$log_passengers <- log1p(airline_data_cleaned$passengers)  # log1p handles zero values

# Plot the distribution of log-transformed passengers
ggplot(airline_data_cleaned, aes(x = log_passengers)) +
  geom_histogram(bins = 30, fill = "green", color = "black") +
  labs(
    title = "Distribution of Log-Transformed Passengers",
    x = "Log-Transformed Passengers",
    y = "Frequency"
  ) +
  theme_minimal()

# Apply log transformation to fares
airline_data_cleaned$log_fare <- log1p(airline_data_cleaned$fare)

# Plot the distribution of log-transformed fares
ggplot(airline_data_cleaned, aes(x = log_fare)) +
  geom_histogram(bins = 30, fill = "orange", color = "black") +
  labs(
    title = "Distribution of Log-Transformed Fares",
    x = "Log-Transformed Fare",
    y = "Frequency"
  ) +
  theme_minimal()

```
## Boxplots for Log-Transformed Variables

### Boxplot for Log-Transformed Passengers
```{r boxplot-log-passengers, warning=FALSE, message=FALSE}
# Boxplot for log-transformed passengers
library(ggplot2)

ggplot(airline_data_cleaned, aes(x = log_passengers)) +
  geom_boxplot(fill = "green", color = "black") +
  labs(
    title = "Boxplot of Log-Transformed Passengers",
    x = "Log-Transformed Passengers"
  ) +
  theme_minimal()

# Boxplot for log-transformed fares
ggplot(airline_data_cleaned, aes(x = log_fare)) +
  geom_boxplot(fill = "orange", color = "black") +
  labs(
    title = "Boxplot of Log-Transformed Fares",
    x = "Log-Transformed Fares"
  ) +
  theme_minimal()
```
## Aggregations, Correlations, and Pairwise Relationships

### 1. Aggregate and Plot Average Log Passengers Over Time
```{r log-passengers-over-time, warning=FALSE, message=FALSE}
# Aggregate log_passengers by time
library(dplyr)
library(ggplot2)

average_passengers_over_time <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_log_passengers = mean(log_passengers, na.rm = TRUE)) %>%
  ungroup()

# Plot the aggregated time series
ggplot(average_passengers_over_time, aes(x = time, y = average_log_passengers)) +
  geom_line(color = "green", size = 1) +
  labs(
    title = "Average Log Passengers Over Time",
    x = "Time",
    y = "Average Log Passengers"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey90"))

# Aggregate log_fare by time
average_fare_over_time <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_log_fare = mean(log_fare, na.rm = TRUE)) %>%
  ungroup()

# Plot the aggregated time series
ggplot(average_fare_over_time, aes(x = time, y = average_log_fare)) +
  geom_line(color = "orange", size = 1) +
  labs(
    title = "Average Log Fare Over Time",
    x = "Time",
    y = "Average Log Fare"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey90"))

# Aggregate log_fare by time
average_fare_over_time <- airline_data_cleaned %>%
  group_by(time) %>%
  summarise(average_log_fare = mean(log_fare, na.rm = TRUE)) %>%
  ungroup()

# Plot the aggregated time series
ggplot(average_fare_over_time, aes(x = time, y = average_log_fare)) +
  geom_line(color = "orange", size = 1) +
  labs(
    title = "Average Log Fare Over Time",
    x = "Time",
    y = "Average Log Fare"
  ) +
  theme_minimal() +
  theme(panel.grid.major = element_line(color = "grey90"))

# Create pairwise scatterplots
library(GGally)

ggpairs(
  airline_data_cleaned,
  columns = c("log_fare", "log_passengers", "nsmiles"),
  diag = list(continuous = "densityDiag"),
  lower = list(continuous = "smooth"),
  upper = list(continuous = "cor")
) +
  labs(title = "Pairwise Relationships Among Key Variables") +
  theme_minimal()

```

